---
title: "Presentation"
author: "Virginia Baskin"
date: "4/13/2023"
output: ioslides_presentation
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
source("shiny_plot_setup.R", echo=FALSE)
#show_col_types = FALSE
```

## Shiny Presentation

This R Markdown presentation is made interactive using Shiny. The viewers of the presentation can change the assumptions underlying what's presented and see the results immediately. 

To learn more, see [Interactive Documents](http://rmarkdown.rstudio.com/authoring_shiny.html).

## Interactive Plot

```{r eruptions}
inputPanel(
  selectInput("n_breaks", label = "Number of bins:",
              choices = c(10, 20, 35, 50), selected = 20),
  
  sliderInput("bw_adjust", label = "Bandwidth adjustment:",
              min = 0.2, max = 2, value = 1, step = 0.2)
)

renderPlot({
  hist(faithful$eruptions, probability = TRUE, breaks = as.numeric(input$n_breaks),
       xlab = "Duration (minutes)", main = "Geyser eruption duration")
  
  dens <- density(faithful$eruptions, adjust = input$bw_adjust)
  lines(dens, col = "blue")
})
```

## Bullets

- Bullet 1
- Bullet 2
- Bullet 3

## R Output

```{r cars}
summary(cars)
```

## Data 
- Problem and Importance

## Data Introduction 
- bullet points 
- bullet points 
- bullet points 
- dude our data was all categorical so when i tell u it was difficult to plot


## Supplementary Datasets


## Exploration 

- plot of citations by race over time  
- plot of frequency of citation versus hour in day  
- mph over speed limit by year  

## Pyramid 
```{r, message = FALSE}
source("subscripts/citation_pyramid.R", local = TRUE, print.eval = FALSE, echo= FALSE, verbose= FALSE)
pyramid_cites()
```


## Violin
```{r, message = FALSE}
source("subscripts/violin_speed.R")
violin_speed()
```


--- 
## Killer Plot 

## Shiny

---

#Race Line Graph by Year: line_ry()
#waffle plot: waffle()



```{r, echo=FALSE}

# black
black <- mini_htx[mini_htx$subject_race=='black',]
blk_total <- nrow(black)

blk_fin <- black %>% filter(violation=="FAILURE TO ESTABLISH FINANCIAL RESPONSIBILITY") %>%
         summarise(n_individuals = n()) %>%
         arrange(desc(n_individuals))
blk_perc <- blk_fin[1]/blk_total

# white
white <- mini_htx[mini_htx$subject_race=='white',]
white_total <- nrow(white)

white_fin <- white %>% filter(violation=="FAILURE TO ESTABLISH FINANCIAL RESPONSIBILITY") %>%
         summarise(n_individuals = n()) %>%
         arrange(desc(n_individuals))
white_perc <- white_fin[1]/white_total

# aapi
aapi <- mini_htx[mini_htx$subject_race=='asian/pacific islander',]
aapi_total <- nrow(aapi)

aapi_fin <- aapi %>% filter(violation=="FAILURE TO ESTABLISH FINANCIAL RESPONSIBILITY") %>%
         summarise(n_individuals = n()) %>%
         arrange(desc(n_individuals))
aapi_perc <- aapi_fin[1]/aapi_total

# unknown
unknown <- mini_htx[mini_htx$subject_race=='unknown',]
unknown_total <- nrow(unknown)

unknown_fin <- unknown %>% filter(violation=="FAILURE TO ESTABLISH FINANCIAL RESPONSIBILITY") %>%
         summarise(n_individuals = n()) %>%
         arrange(desc(n_individuals))
unknown_perc <- unknown_fin[1]/unknown_total

# NA
nas <- mini_htx[is.na(mini_htx$subject_race),]
nas_total <- nrow(unknown)

nas_fin <- nas %>% filter(violation=="FAILURE TO ESTABLISH FINANCIAL RESPONSIBILITY") %>%
         summarise(n_individuals = n()) %>%
         arrange(desc(n_individuals))
nas_perc <- nas_fin[1]/nas_total

all_perc <- data.frame(c(blk_perc, white_perc, aapi_perc, unknown_perc, nas_perc))
colnames(all_perc) <- c("black", "white", "aapi", "unknown","N/A")
all_perc
```
Of all recorded races, the proportion of total citations that mention failure to establish financial responsibility as the sole violation is the highest for black individuals. This proportion is about 0.01 higher for black individuals compared to white individuals, while aapi individuals and those of unknown race share a similar proportion around 0.0025. However, the greatest proportion of total citations that mention failure to establish financial responsibility as the sole violation is attributed to "N/A", which indicates that the officer neglected or failed to record the subject's race. 

```{r hypothesis test}

hyptest_finresp <- function(race1, race2, sig){
  pval <- prop.test(x = c(pull(get(paste0(race1, "_fin"))), pull(get(paste0(race2, "_fin"))))
                    , n = c(get(paste0(race1, "_total")),
                            get(paste0(race2,"_total"))), alternative='greater')$p.value
  if (pval<sig){
    cat("pval is:", pval, "\nThe null hypothesis is rejected. \n")
  }
  if (pval>=sig){
    cat("pval is:", pval, "\nThe null hypothesis is not rejected. \n")
  }
}

for (race in c("white", "aapi", "unknown", "nas")){
  hyptest_finresp("blk", race, 0.05)
}

for (race in c("aapi", "unknown")){
  hyptest_finresp("white", race, 0.05)
}

```





---
title: "Final Presentation"
author: "Siddhi Narayan, Grace Wang, Virginia Baskin"
date: "2023-04-17"
output: 
  ioslides_presentation:
    smaller: true
    widescreen: true
    css: style.css
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.width = 9, fig.height = 5, out.width = "100%")
source("shiny_plot_setup.R", echo=FALSE)
source("Tables/Racial_Citation.R", echo = FALSE)
source("Tables/Racial_Breakdown.R", echo = FALSE)
source("Tables/FinResp.R", echo = FALSE)
source("Tables/average_speeding_amt.R", echo = FALSE)
source("Tables/csqm.R", echo = FALSE)
```

# Introduction

## Primary Dataset: Citations in HTX
Insert description

## Secondary Datasets
* Census Data
* GIS/Beats Data


## Exploration
- bullet 
- point

# Distribution Over Time

## Citation Count by Race over the Years {.smaller}
```{r Race/Year Line, message=FALSE}
source("subscripts/line_race_year.R", local=TRUE, print.eval=TRUE)
line_ry()
```

## Citation Frequency for Time of Day {.smaller}
```{r Time of Day Freq, message=FALSE}
source("subscripts/line_timeofcites.R", local=TRUE, print.eval=TRUE)
time_line()
```

# Distribution by Demographic

## Citation by Race {.smaller}
```{r Waffle, message=FALSE}
source("subscripts/waffle.R", local=TRUE, print.eval=TRUE)
waffle()
```

## MPH over Speed Limit for Race & Gender {.smaller}
```{r Violin, message=FALSE}
source("subscripts/violin_speed.R", local=TRUE, print.eval=TRUE)
violin_speed()
```

# Distribution by Location

## Citation by Sex by District {.smaller}
```{r Pyramid, message = FALSE}
source("subscripts/citation_pyramid.R", local = TRUE, print.eval = TRUE)
pyramid_cites()
```

## Citations Per Square Mile | Top 5 Beats ranked by citations/square mile:
```{r csqm_top, echo = FALSE, message = FALSE}
knitr::kable(csqm_top)
```

## Citations Per Square Mile | Bottom 5 Beats ranked by citations/square mile
```{r csqm_bottom, echo = FALSE, message = FALSE}
knitr::kable(csqm_bottom)
```

## Citations Per Square Mile | Statistics on citations per square mile:
```{r csqm_summry, echo = FALSE, message = FALSE}
knitr::kable(cqm_summary)
```

## Killer Plot
<iframe src="http://127.0.0.1:5501" width="100%" height="500px"></iframe>

# Analysis

## Houston Racial Breakdown vs Citations Racial Breakdown {.smaller}
```{r racial_citation_breakdown, echo = FALSE, message = FALSE}
knitr::kable(htx)
```

## Citation Type Racial Breakdown {.smaller}
Speeding Citations by Race:
```{r speeding_race, echo = FALSE, message = FALSE}
knitr::kable(speeding_race)
```
##
Invalid License Citations by Race:
```{r license_race, echo = FALSE, message = FALSE}
knitr::kable(license_race)
```

##
Failure to Establish Financial Responsibility Citations by Race:
```{r financial_race, echo = FALSE, message = FALSE}
knitr::kable(financial_race)
```

##
Seat Belt Citations by Race:
```{r seat_belt_race, echo = FALSE, message = FALSE}

knitr::kable(seat_belt_race)
```

##
Running a Red Light/Stop Sign Citations by Race:
```{r stop_light_race, message = FALSE}
knitr::kable(stop_light_race)
```

## Failure to Establish Financial Responsibility {.smaller}
```{r fin_resp, echo = FALSE, message = FALSE}
knitr::kable(all_perc)
```


## Average Speeding Amounts {.smaller}
All People (baseline):
```{r all_speed, echo = FALSE, message = FALSE}
knitr::kable(all_speed)
```
##
By Race:
```{r race_speed, echo = FALSE, message = FALSE}
knitr::kable(race_speed)
```
##
By Gender:
```{r gender_speed, echo = FALSE, message = FALSE}
knitr::kable(gender_speed)
```
##
```{r race_gender_speed, echo = FALSE, message = FALSE}
library(kableExtra)
kable(race_gender_speed) %>% kable_styling(font_size=8)
``` 
# Conclusion

## Conclusions
conclusions



Speeding Citations by Race:
```{r speeding_race, echo = FALSE, message = FALSE}
knitr::kable(speeding_race)
```
##
Invalid License Citations by Race:
```{r license_race, echo = FALSE, message = FALSE}
knitr::kable(license_race)
```

##
Failure to Establish Financial Responsibility Citations by Race:
```{r financial_race, echo = FALSE, message = FALSE}
knitr::kable(financial_race)
```

##
Seat Belt Citations by Race:
```{r seat_belt_race, echo = FALSE, message = FALSE}

knitr::kable(seat_belt_race)
```

##
Running a Red Light/Stop Sign Citations by Race:
```{r stop_light_race, message = FALSE}
knitr::kable(stop_light_race)
```

